Towards Practical First-Order Model Counting

### 2025-01-16 doc/paper/paper.tex
Add AAAI feedback
Move TODOs around
Rewrite inline enumerates

%\acknowledgements{I want to thank \dots}%optional

@@

@@

\section{Introduction}

@@

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime
approximation~\cite{DBLP:conf/ijcai/BremenK20}. Among these, FOKC-based
algorithms are particularly prominent, transforming formulas into structured
representations such as circuits or graphs. Notable examples include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its successor
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. Another important algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, uses cell enumeration as its
foundation.

@@

The \textsc{Crane} algorithm marked a significant step forward, expanding the
range of formulas handled by FOMC algorithms. However, it had notable
limitations:
\begin{enumerate}
  \item it required manual evaluation of function definitions to compute model
  counts, and
  \item it introduced recursive functions without proper base cases, making it
  more complex to use.
\end{enumerate}
To address these shortcomings, we present \Cranetwo{}, a fully automated FOMC
algorithm that overcomes the constraints of its predecessor. \Cranetwo{} can
handle domain sizes over \num{500000} times larger than previous algorithms and
simplifies the user experience by automatically handling base cases and
compiling function definitions into efficient C++ programs.

@@

@@

In this section, we will review the basic concepts of first-order logic as they
are used in FOKC algorithms. There are two key differences between the logic
used by these algorithms and the logic supported as input. First,
Skolemization~\cite{DBLP:conf/kr/BroeckMD14} eliminates existential quantifiers
by introducing additional predicates. Please note that Skolemization here
differs from the standard Skolemization procedure that introduces function
symbols~\cite{DBLP:books/daglib/0030198}. Second, the input formula is rewritten
as a conjunction of clauses, each in \emph{prenex normal
  form}~\cite{hinman2018fundamentals}.

A \emph{term} can be either a variable or a constant. An \emph{atom} can be
either
\begin{enumerate}
  \item $P(t_{1}, \dots, t_{m})$ for some predicate $P$ and terms
  $t_{1}, \dots, t_{m}$ (written as $P(\mathbf{t})$ for short) or
  \item $x=y$ for some terms $x$ and $y$.
\end{enumerate}
The \emph{arity} of a predicate is the number of arguments it takes, i.e., $m$
in the case of the predicate $P$ mentioned above. We write $P/m$ to denote a
predicate along with its arity. A \emph{literal} can be either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
}\forall x_{2} \in \Delta_{2}\dots\text{ }\forall x_{n} \in \Delta_{n}\text{.
}\phi(x_{1}, x_{2}, \dots, x_{n})$, where $\phi$ is a disjunction of literals
that only contain variables $x_{1}, \dots, x_{n}$ (and any constants). We say
that a clause is a \emph{(positive) unit clause} if
\begin{enumerate}
  \item there is only one literal with a predicate, and
  \item it is a positive literal.
\end{enumerate}
Finally, a \emph{formula} is a conjunction of clauses. Throughout the paper, we
will use set-theoretic notation, interpreting a formula as a set of clauses and
a clause as a set of literals.

@@

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption[]{A comparison of the three logics used in FOMC based on the following aspects:
    \begin{enumerate}
      \item the number of sorts,
      \item support for constants,
      \item the maximum number of variables,
      \item supported quantifiers, and
      \item supported atoms in addition to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple of terms $\mathbf{t}$.
    \end{enumerate}
    Here:
    \begin{enumerate}
      \item $k$ and $m$ are non-negative integers, with the latter depending on the domain size,
      \item $P$ represents a predicate, and
      \item $x$ and $y$ are terms.
    \end{enumerate}
  }\label{tbl:logics}
\end{table*}

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in FOMC:
\begin{enumerate}
  \item \FO{} is the input format for
  \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
  its extensions
  \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
  \Cranetwo{};
  \item \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
  related
  methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22};
  \item \UFO{} is the input format supported by the most recent implementation
  of
  \textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}}.
\end{enumerate}
The notation we use to refer to each logic is standard in the case of \Ctwo{}
and \UFO{}~\cite{tóth2024complexityweightedfirstordermodel} and redefined to be
more specific in the case of \FO{}. All three logics are function-free, and
domains are always assumed to be finite. As usual, we presuppose the
\emph{unique name assumption}, which states that two constants are equal if and
only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

@@

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for
\begin{enumerate}
  \item constants,
  \item the equality predicate,
  \item multiple domains, and
  \item formulas with more than two variables.
\end{enumerate}
The advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains. Let
  $\sigma$ be a map from the domains of $\phi$ to their interpretations as sets,
  satisfying the following conditions:
  \begin{enumerate}
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{enumerate}
  A \emph{structure} of $\phi$ is a set $M$ of ground literals defined by adding
  to $M$ either $P(\mathbf{t})$ or $\neg P(\mathbf{t})$ for every predicate
  $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}

@@

While this work focuses on FOMC, we still define the weighted variant of the
problem as Skolemization relies on weights even for unweighted FOMC.

\begin{definition}[WFOMC instance]\label{def:instance}
  A \emph{WFOMC instance} comprises:
  \begin{enumerate}
    \item a formula $\phi$ in \FO{},
    \item two (rational) \emph{weights} $w^{+}(P)$ and $w^{-}(P)$ assigned to
    each predicate $P$ in $\phi$, and
    \item $\sigma$ as described in \cref{def:model}.
  \end{enumerate}
  Unless specified otherwise, we assume all weights to be equal to 1.
\end{definition}

@@

@@

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a formula $\phi$. First, the formula is compiled into a set of equations, which are then used to create a C++ program. This program can be executed with different command line arguments to calculate the model count of $\phi$ for different domain sizes. To accomplish this, the \CompileWithBaseCases function employs several components:
    \begin{enumerate}
      \item the FOKC algorithm of \textsc{Crane},
      \item a procedure called \FindBaseCases, which identifies a sufficient set of base cases,
      \item a procedure called \Propagate, which constructs a formula corresponding to a given base case, and
      \item algebraic simplification techniques (denoted as \Simplify).
    \end{enumerate}
  }\label{fig:overview}
\end{figure*}

@@

Before describing the main contribution of this work, let us review the
essential aspects of FOKC as realised by \textsc{Crane}. The input formula is
compiled into:
\begin{enumerate}
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate}
$\mathcal{E}$ can contain any number of functions, one of which (denoted by $f$)
represents the solution to the FOMC problem. To compute the FOMC for particular
domain sizes, $f$ must be evaluated with those domain sizes as arguments.
$\mathcal{D}$ records this correspondence between function arguments and
domains.

@@

\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

@@

Finally, the main responsibility of the \Simplify procedure is to handle the
algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here:
\begin{enumerate}
  \item $n$ is a variable,
  \item $a, b \in \mathbb{N}_{0}$ are constants,
  \item $f$ is an expression that may depend on $m$, and
  \item $[a \le m \le b] =
  \begin{cases}
    1 & \text{if $a \le m \le b$} \\
    0 & \text{otherwise}
  \end{cases}$.
\end{enumerate}
\Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overall}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
$g(l, m) = g(l-1, m) + mg(l-1, m-1)$.

\subsection{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}

@@

\Cref{alg:findbasecases} summarises the implementation of \FindBaseCases.
\FindBaseCases considers two types of arguments when a function $f$ calls itself
recursively:
\begin{enumerate}
  \item constants and
  \item arguments of the form $x_{i} - c_{i}$, where $c_{i}$ is a constant and
  $x_{i}$ is the $i$-th argument of the signature of $f$.
\end{enumerate}
When the argument is a constant $c_{i}$, a base case with $c_{i}$ is added. In
the second case, a base case is added for each constant from $0$ up to (but not
including) $c_{i}$.

@@

In the case of $n = 0$, there are three types of clauses to consider:
\begin{enumerate}
  \item those that do not mention $\Delta$,\label[type]{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$, and\label[type]{type2}
  \item those that have some literals with variables quantified over $\Delta$
  and some without.\label[type]{type3}
\end{enumerate}
Clauses of \cref{type1} are transferred to the new formula $\phi'$ without any
changes. For clauses of \cref{type2}, $C'$ is empty, so these clauses are
filtered out. As for clauses of \cref{type3}, a new kind of smoothing is
performed, which will be explained in \cref{sec:smoothing}.

@@

\subsection{Generating C++ Code}\label{sec:cpp}

@@

@@

\subsection{Benchmarks}
We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by Dilkas and Belle~\cite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' Markov
logic network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In
\Ctwo{}, \FO{}, and \UFO{}, this problem can be formulated as
\[
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\]
or, equivalently, in conjunctive normal form as
\[
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\]
Finally, we include the bijection-counting problem previously utilised by Dilkas
and Belle~\cite{DBLP:conf/kr/DilkasB23}. Its formulation in \FO{} is described
in \cref{example:overall}. The equivalent formula in \Ctwo{} is
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{. }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{. }P(x, y)).
\]
Similarly, in \UFO{} the same formula can be written as
\[
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.

\begin{figure*}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The runtime of the algorithms as a function of the domain size. Note
    that both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure*}
% TODO: shrink to fit into the margins

@@

\subsection{Results}
\Cref{fig:plot} presents a summary of the experimental results. Only
\textsc{FastWFOMC} and \Cranebfs{} could handle the bijection-counting problem.
For this benchmark, the largest domain sizes these algorithms could accommodate
were \num{64} and \num{4096}, respectively. On the other two benchmarks,
\textsc{ForcLift} had the lowest runtime. However, due to its finite precision,
it only scaled up to domain sizes of \num{16} and \num{128} for `friends' and
`functions', respectively. \textsc{FastWFOMC} outperformed \textsc{ForcLift} in
the case of `friends', but not `functions', as it could handle domains of size
\num{1024} and \num{64}, respectively. Furthermore, both \Cranebfs{} and
\Cranegreedy{} performed similarly on both benchmarks. Similarly to the
`bijections' benchmark, \Cranetwo{} significantly outperformed the other two
algorithms, scaling up to domains of size \num{8192} and \num{67108864},
respectively.

>>>>

\acknowledgements{Part of the research was conducted while all authors were at the National University of Singapore.}

@@

% TODO: the technical appendix should follow the same template (probably just
% integrate it into the main paper)

% TODO (for the entire paper):
% * 15 pages excluding references
% * draw a clear line between prior work and new developments

@@

\section{Introduction}
% TODO: We would like to clarify that the main contribution of this work
% consists of everything needed to complement recursive function definitions
% with the necessary base cases. This process includes identifying the base
% cases and their corresponding formulas, transforming them (including applying
% a new smoothing procedure), and recursively reusing Crane. C++ code
% generation, although relatively straightforward, is crucial for the usability
% of the algorithm

% TODO: focus more on current work

@@

% TODO: don't assume that the reader knows FOMC and FOMC algorithms

% TODO: It would be nice to give the necessary introduction to FOKC, or CRANE
% (an intuitive explanation early in the paper)

% TODO: The principal motivation of a knowledge compilation-based approach to
% FOMC comes from its ability to find polynomial-time solutions, where the
% polynomial has a lower degree compared to, e.g., FastWFOMC (Dilkas and Belle,
% 2023). The ability of the algorithm to mathematically describe recursive
% solutions is also noteworthy as the only other available alternative is to
% guess them (Barvı́nek, van Bremen, Wang, Zelezný, Kuzelka, ILP 2021).
FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}, and anytime
approximation~\cite{DBLP:conf/ijcai/BremenK20}. Among these, FOKC-based
algorithms are particularly prominent, transforming formulas into structured
representations such as circuits or graphs. Notable examples include
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its successor
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}. Another important algorithm,
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}, uses cell enumeration as its
foundation.

@@

The \textsc{Crane} algorithm marked a significant step forward, expanding the
range of formulas handled by FOMC algorithms. However, it had notable
limitations: it required manual evaluation of function definitions to compute
model counts and introduced recursive functions without proper base cases,
making it more complex to use. To address these shortcomings, we present
\Cranetwo{}, a fully automated FOMC algorithm that overcomes the constraints of
its predecessor. \Cranetwo{} can handle domain sizes over \num{500000} times
larger than previous algorithms and simplifies the user experience by
automatically handling base cases and compiling function definitions into
efficient C++ programs.

@@

% TODO: focus more on current work

% TODO: Section 2 until 2.3 (Algebra) could be drastically shortened

@@

% TODO: Skolemization is a procedure that transforms a formula with existential
% quantifiers into an ’equivalent’ formula without existential quantifiers with
% the same WFOMC.
In this section, we will review the basic concepts of first-order logic as they
are used in FOKC algorithms. There are two key differences between the logic
used by these algorithms and the logic supported as input. First,
Skolemization~\cite{DBLP:conf/kr/BroeckMD14} eliminates existential quantifiers
by introducing additional predicates. Please note that Skolemization here
differs from the standard Skolemization procedure that introduces function
symbols~\cite{DBLP:books/daglib/0030198}. Second, the input formula is rewritten
as a conjunction of clauses, each in \emph{prenex normal
  form}~\cite{hinman2018fundamentals}.

A \emph{term} can be either a variable or a constant. An \emph{atom} can be
either $P(t_{1}, \dots, t_{m})$ (or $P(\mathbf{t})$) for some predicate $P$ and
terms $t_{1}, \dots, t_{m}$ or $x=y$ for some terms $x$ and $y$. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of the predicate $P$ mentioned above. We write $P/m$ to denote a
predicate along with its arity. A \emph{literal} can be either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
}\forall x_{2} \in \Delta_{2}\dots\text{ }\forall x_{n} \in \Delta_{n}\text{.
}\phi(x_{1}, x_{2}, \dots, x_{n})$, where $\phi$ is a disjunction of literals
that only contain variables $x_{1}, \dots, x_{n}$ (and any constants). We say
that a clause is a \emph{(positive) unit clause} if there is only one literal
with a predicate, and it is a positive literal. Finally, a \emph{formula} is a
conjunction of clauses. Throughout the paper, we will use set-theoretic
notation, interpreting a formula as a set of clauses and a clause as a set of
literals.

% TODO: Page 2, Section 2.1, paragraph 2, line 10 - ”A clause is of the form ...” - are the quantifiers really part of the clause? Shouldn’t it be possible to quantify a conjunction of several clauses?
% Conforming with previous work, the definition of a clause includes universal quantifiers for all variables within. While it is possible to rewrite the entire formula in prenex normal form, the format we describe has proven itself convenient to work with.

@@

% TODO: shorten the part about the three logics

\begin{table*}[t]
  \centering
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Additional atoms\\
    \midrule
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & one & \xmark & two & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & one & \xmark & two & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
  \caption{A comparison of the three logics used in FOMC\@. The
    2\textsuperscript{nd}--5\textsuperscript{th} columns refer to: the number of
    sorts, support for constants, the maximum number of variables, and supported
    quantifiers, respectively. The last column lists supported atoms in addition
    to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple
    of terms $\mathbf{t}$. Here: $k$ and $m$ are non-negative integers, with the
    latter depending on the domain size, $P$ represents a predicate, and $x$ and
    $y$ are terms. }\label{tbl:logics}
\end{table*}

@@

In \cref{tbl:logics}, we outline the differences among three first-order logics
commonly used in FOMC: \FO{}, \Ctwo{}, and \UFO{}. First, \FO{} is the input
format for
\textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}} and
its extensions
\textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}} and
\Cranetwo{}. Second, \Ctwo{} is often used in the literature on
\textsc{FastWFOMC} and related
methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. Finally,
\UFO{} is the input format supported by the most recent implementation of
\textsc{FastWFOMC}\footnote{\url{https://github.com/jan-toth/FastWFOMC.jl}}. The
notation we use to refer to each logic is standard in the case of \Ctwo{} and
\UFO{}~\cite{tóth2024complexityweightedfirstordermodel} and redefined to be more
specific in the case of \FO{}. All three logics are function-free, and domains
are always assumed to be finite. As usual, we presuppose the \emph{unique name
  assumption}, which states that two constants are equal if and only if they are
the same constant~\cite{DBLP:books/aw/RN2020}.

@@

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for constants, the equality
predicate, multiple domains, and formulas with more than two variables. The
advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains. Let
  $\sigma$ be a map from the domains of $\phi$ to their interpretations as sets
  such that the sets are pairwise disjoint, and the constants in $\phi$ are
  included in the corresponding domains. A \emph{structure} of $\phi$ is a set
  $M$ of ground literals defined by adding to $M$ either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. A structure is a
  \emph{model} if it satisfies $\phi$.
\end{definition}
% TODO: Page 2, Definition 1 - ”A structure is a model if it satisfies ϕ” - do you mean that it makes ϕ valid?

@@

While this work focuses on FOMC, we still define the weighted variant of the
problem as Skolemization relies on weights even for unweighted FOMC.
% TODO: Even though we are evaluating the algorithms on FOMC benchmarks, for
% sentences with existential quantifiers, computing FOMC using Gantry requires
% the use of WFOMC. For such sentences, Skolemization introduces predicates with
% non-unary weights that must be accounted for to compute the correct model
% count.

\begin{definition}[WFOMC instance]\label{def:instance}
  A \emph{WFOMC instance} comprises: a formula $\phi$ in \FO{}, two (rational)
  \emph{weights} $w^{+}(P)$ and $w^{-}(P)$ assigned to each predicate $P$ in
  $\phi$, and $\sigma$ as described in \cref{def:model}. Unless specified
  otherwise, we assume all weights to be equal to 1.
\end{definition}

@@

% TODO: this subsection could be expanded a little

@@

% TODO: The algorithm’s recursive nature and its interaction with CRANE could be
% explained more clearly

% TODO: more emphasis on the motivation for each part of GANTRY (e.g., how
% FindBaseCases addresses key issues in CRANE) would add depth to the
% contribution.

% TODO: Section 3: you seem to use "equation" and "function" interchangeably

% TODO: this section "would benefit from cleaner reasoning"

% TODO: this section should be expanded by 1 page ("a clear discussion of the
% innovation aspect and technical parts")

@@

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a
    formula $\phi$. First, the formula is compiled into a set of equations,
    which are then used to create a C++ program. This program can be executed
    with different command line arguments to calculate the model count of $\phi$
    for different domain sizes. To accomplish this, the \CompileWithBaseCases
    function makes use of the FOKC algorithm of \textsc{Crane}, algebraic
    simplification techniques (denoted as \Simplify), and two other auxiliary
    procedures. }\label{fig:overview}
\end{figure*}

@@

Before describing the main contribution of this work, let us review the
essential aspects of FOKC as realised by \textsc{Crane}. The input formula is
compiled into: set $\mathcal{E}$ of equations, map $\mathcal{F}$ from function
names to formulas, and map $\mathcal{D}$ from function names and argument
indices to domains. $\mathcal{E}$ can contain any number of functions, one of
which (denoted by $f$) represents the solution to the FOMC problem. To compute
the FOMC for particular domain sizes, $f$ must be evaluated with those domain
sizes as arguments. $\mathcal{D}$ records this correspondence between function
arguments and domains.

@@

% TODO: move this example right after the definition of WFOMC (or near the
% definition of a model)
\begin{example}[Counting bijections]\label{example:overall}
  Consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \]
  We specifically examine the first solution returned by \Cranetwo{} for this
  formula.

@@

Finally, the main responsibility of the \Simplify procedure is to handle the
algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here: $n$ is a variable,
$a, b \in \mathbb{N}_{0}$ are constants, and $f$ is an expression that may
depend on $m$. Additionally, $[a \le m \le b] = \begin{cases}
  1 & \text{if $a \le m \le b$} \\
  0 & \text{otherwise}
\end{cases}$.
\Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overall}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
$g(l, m) = g(l-1, m) + mg(l-1, m-1)$.

\subsection{Identifying a Sufficient Set of Base Cases}\label{sec:identifying}
% TODO: prove correctness?

@@

\Cref{alg:findbasecases} summarises the implementation of \FindBaseCases. It
considers two types of arguments when a function $f$ calls itself recursively:
constants and arguments of the form $x_{i} - c_{i}$. Here, $c_{i}$ is a
constant, and $x_{i}$ is the $i$-th argument of the signature of $f$. When the
argument is a constant $c_{i}$, a base case with $c_{i}$ is added. In the second
case, a base case is added for each constant from $0$ up to (but not including)
$c_{i}$.

@@

% NOTE: keep this enumerate
In the case of $n = 0$, there are three types of clauses to consider:
\begin{enumerate}
  \item those that do not mention $\Delta$,\label[type]{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$, and\label[type]{type2}
  \item those that have some literals with variables quantified over $\Delta$
  and some without.\label[type]{type3}
\end{enumerate}
Clauses of \cref{type1} are transferred to the new formula $\phi'$ without any
changes. For clauses of \cref{type2}, $C'$ is empty, so these clauses are
filtered out. As for clauses of \cref{type3}, a new kind of smoothing is
performed, which will be explained in \cref{sec:smoothing}.

@@

\subsection{Generating C++ Code}\label{sec:cpp}
% TODO (must): expand this section
% TODO: The purpose of C++ code generation is to turn mathematical equations into executable code.

@@

% TODO Experiments could be expanded by 1 page ("a more thorough and independent
% practical assessment")

% TODO: this section "would benefit from cleaner reasoning"

% TODO: The difference between Gantry-Greedy and Gantry-BFS is in how the
% algorithm chooses which compilation rule to apply to a formula. The former
% uses greedy search: there is a list of rules, and the first applicable rule is
% the one that gets used, disregarding all the others. The latter uses a
% combination of greedy and breadth-first search (BFS). That is, each
% compilation rule is identified as either greedy or non-greedy. Greedy rules
% are applied as soon as possible at any stage of the compilation process. BFS
% is executed over all applicable non-greedy rules, identifying the solution
% that can be constructed using the smallest number of such rules.

@@

\subsection{Benchmarks}
% TODO: more benchmarks from NOT my work
% TODO: make sure it's clear what the 'raw' instances are
% TODO: don't emphasise that the benchmarks come from my previous work

We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by Dilkas and Belle~\cite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' Markov
logic network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In
\Ctwo{}, \FO{}, and \UFO{}, this problem can be formulated as
\[
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\]
or, equivalently, in conjunctive normal form as
\[
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\]
Finally, we include the bijection-counting problem previously utilised by Dilkas
and Belle~\cite{DBLP:conf/kr/DilkasB23}. Its formulation in \FO{} is described
in \cref{example:overall}. The equivalent formula in \Ctwo{} is
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{. }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{. }P(x, y)).
\]
Similarly, in \UFO{} the same formula can be written as
\[
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.

\begin{figure*}[t]
  \centering
  \includegraphics{plot.pdf}
  \caption{The runtime of the algorithms as a function of the domain size. Note
    that both axes are on a logarithmic scale.}\label{fig:plot}
\end{figure*}
% TODO: shrink/restructure to fit into the margins

@@

\subsection{Results}

% TODO: On the ’friends’ and ’functions’ benchmarks, ForcLift runs until the
% model count exceeds 231 − 1.

% TODO: We are not aware of any formulas on which Gantry scales worse compared
% to either ForcLift or FastWFOMC. The one advantage that FastWFOMC has over
% Gantry is its support for counting quantifiers.

% TODO: Regarding programming languages and accuracy, we verified that the
% answers match for smaller domain sizes. Also, although written in different
% languages, both Gantry and FastWFOMC use the GNU Multiple Precision Arithmetic
% Library.
